---
title: "PreclinAD_Microbiome_Analyses"
output: html_document
date: "2025-08-15"
---

```{r setup, include=FALSE}
library(ggplot2)          #3.5.0
library(ggExtra)          #0.10.1
library(ggridges)
library(ggrepel)
library(dplyr)            #1.1.2
library(phyloseq)         #1.46.0
library(vegan)
library(Maaslin2)         #1.12.0
library(ape)              #5.7.1
library(readxl)           #1.4.3     
library(microViz)         #0.12.0
library(ggside)           #0.3.1
library(patchwork)        #1.2
library(ComplexHeatmap)   #2.14.0
library(tidyr)            #1.3.1
library(readr)            #2.1.5
library(reshape2)         #1.4.4
library(rstatix)          #0.7.2
library(ggpubr)           #0.6.0
library(NatParksPalettes) #0.2.0
library(cowplot)          #1.1.2
library(stringr)          #1.5.1
library(Boruta)           #8.0.0 
library(caret)            #6.0-94
library(VIM)              #6.2.2 
library(corpcor)
library(corrplot)
library(Hmisc)
library(ANCOMBC)          #2_4.0   
library(MTXmodel)         #1.2.4
library(forcats)          #1.0.0
library(tidyverse)        #2.0.0  
library(scales)           #1.3.0
library(microbiome)
library(lme4)
library(lmerTest)
library(lmtest)
library(broom)
library(MuMIn)
library(car)
library(emmeans)
library(igraph)
library(FSA)
library(rcompanion)
## Data Processing Functions 
source("Functions/phyloseq_object.R")
source("Functions/tree_object.R")
source("Functions/data_parsing.R")
source("Functions/Correlation_analyses.R")
source("Functions/GLM_functions.R")
#source("main/HighstatLib.r")

##Analysis and Plotting Functions 
source("Functions/oordinate_wrapper.R")
source("Functions/ordinate_metaphlan.R")
source("Functions/maaslin_MTX_DE.R")
source("Functions/plot_maaslin_MTX_DE.R")
source("Functions/compare_longitudinal_beta_diversity.R")
source("Functions/Statistical_tests.R")
source("Functions/secom.R")
source("Functions/SECOM_functions.R")
source("Functions/rf_boruta.R")
source("Functions/customtab.R")
cur.date <- format(Sys.Date(), "%y%m%d")
output_path <- (paste0("outputs/", cur.date, "/", cur.date))
output_path2 <- (paste0("outputs/", cur.date, "/"))


cat_var <- c('Sample_ID','ID','Time_point','Sample_time','Sex','Race',"CDR",'AD_Status', 'Amyloid_positivity', 'Tau_positivity','Tracer_used_for_classification','Tau_marker','amyloid_tau_status','Tracer.mrfree','Tracer.fsuvr','Tracer.av45','Tracer.tau','apoe_genotype','apoe4_carriage','Diabetes_binary',	'Hypertension_binary','Diabetes',	'Hypertension')


num_var <- c("Years_Education","Age","Days_since_CDR",'Days_since_biomarker_assessment',"combined.centiloid","MRFreePET_Centiloid","Days_sinceMrFreePETpib","pib_fsuvr_rsf_tot_cortmean","Centiloid_PiB_fSUVR_rsf_TOT_CORT","Days_sincePETpib","av45_fsuvr_rsf_tot_cortmean","Centiloid_AV45_fSUVR_rsf_TOT_COR","Days_sincePETav45","LUMIPULSE_CSF_AB42","LUMIPULSE_CSF_AB40","LUMIPULSE_CSF_tTau","LUMIPULSE_CSF_pTau","CSF_ab42_ab40_ratio","Days_sinceCSF","Tauopathy","Days_sincePETtau","mr_vol_tot_wmhypointensities","LOAD_CorticalSignature_Thickness","mr_vol_r_hippocampus","mr_vol_l_hippocampus","mr_vol_tot_intracranial","hippocampus_vol_div_icv",	"Days_sinceMRI","NGAL_ug_ml","Calp_ug_ml",	"log2_NGAL","log2_Calp","WEIGHT","HEIGHT","BMI","Calories_kcal",	'Calories_from_fat_kcal',	'Calories_from_SatFat_kcal',	'Protein_g',	'Carbohydrates_g',	'Total_dietary_fiber_g',	'Total_soluble_fiber_g',	'Dietary_fiber_US_2016_g',	'Soluble_fiber_US_2016',	'Total_sugars_g',	'Added_sugar_g','Monosaccharides_g')


# Custom colors for plotting
custom_colors_hp <- c("Healthy" = "#49b9a5", "Preclinical" = "#eebe61")
custom_colors_AD <- c("Healthy" = "#49b9a5", "Preclinical" = "#eebe61", "Symptomatic" = "#a98ec2")
custom_colors_hp_long <- c("Healthy" = "#49b9a5", "Healthy_follow-up" = "#a7d9c1", "Preclinical" = "#eebe61",
                           "Preclinical_follow-up" = "#e9d8b5")
custom_colors_at_status <- c("Healthy A-/T-"="#49b9a5", "Preclinical A+/T-" = "#eebe61", "Preclinical A+/T+" = "#b08d25", "SNAP" =  "#237a80")

```

Load Data
```{r}

meta_data_path <- "~/Documents/GitHub/PreclinicalAD_GutMicrobiome-Inflammation/R/Data/SupplementaryTable1_final.xlsx"
metaphlan_path <- "~/Documents/GitHub/PreclinicalAD_GutMicrobiome-Inflammation/R/Data/MTG_metaphlan3_merged.txt"

## Read in Metaphlan3 tree which is necessary to compute UniFrac distances later ###
m3.tree.Jan19 <- ape::read.tree("~/Documents/GitHub/PreclinicalAD_GutMicrobiome-Inflammation/R/Data/mpa_v30_CHOCOPhlAn_201901_species_tree.nwk.txt")
m3.tree.Jan19$tip.label <- gsub('GCA_[0-9]+\\|', '', m3.tree.Jan19$tip.label)

## Create phyloseq object
ps_MTG <- make_phyloseq_object(metaphlan_path, 
                                 path_to_metadata= meta_data_path, 
                                 tree_object = m3.tree.Jan19, 
                                 cat_vars=cat_var,
                                 num_vars=num_var,
                                 abund_threshold = 0.1, 
                                 compute_mean=FALSE) 

# Subset phyloseq object to the samples closest to the AD biomarker measure 
## N=198
ps_mtg_single_tp <- phyloseq::subset_samples(ps_MTG, Sample_time == "closest")
ps_mtg_single_tp_meta <- data.frame(phyloseq::sample_data(ps_mtg_single_tp))


### Create follow-up dataset by subsetting to last samples from each participant
## Create phyloseq object for last samples (N = 89)
ps_mtg_last_tp_hp <- ps_MTG %>% subset_samples(Sample_time == "last follow-up")
ps_mtg_last_hp_meta <- data.frame(phyloseq::sample_data(ps_mtg_last_tp_hp))

```

Create Table 1 
```{r}
library(kableExtra)
library("anytime") 
library(tibble) # for rownames_to_column() function
library(stringr) # for str_squish()
library(flextable)
library(officer)
library(forcats)
library(tableone)
library(webshot2)

source("~/Documents/GitHub/AD_longitudinal/datasets/Metadata/customtab.R")

meta_closest <- ps_mtg_single_tp_meta
# Set categorical variables as factors 
categorical_vars <- c('Sex','Race','apoe4_carriage','Diabetes','Hypertension','Tau_positivity')

meta_closest[,categorical_vars] <- lapply(meta_closest[,categorical_vars], factor)

numerical_vars <- c('Age','Years_Education','BMI',"combined.centiloid",'Days_since_biomarker_assessment','Calories_kcal','Calories_from_fat_kcal', 'Calories_from_SatFat_kcal','Protein_g','Carbohydrates_g','Total_dietary_fiber_g','Total_soluble_fiber_g','Total_sugars_g','Added_sugar_g','Monosaccharides_g')

meta_closest[,numerical_vars] <- lapply(meta_closest[,numerical_vars], as.numeric)

# Generate metadata comparison table 
comp_vars <- c('Sex','Age','Years_Education','Race','BMI','apoe4_carriage','Diabetes','Hypertension',"combined.centiloid",'Tau_positivity','Days_since_biomarker_assessment','Calories_kcal','Calories_from_fat_kcal', 'Calories_from_SatFat_kcal','Protein_g','Carbohydrates_g','Total_dietary_fiber_g','Total_soluble_fiber_g','Total_sugars_g','Added_sugar_g','Monosaccharides_g')


tab_hp_closest <- CreateTableOne(vars=comp_vars,
                           strata='AD_Status', 
                           data=meta_closest)
tab_plot_hp_closest <- print(tab_hp_closest, 
                        showAllLevels = TRUE,
                        quote = F,
                        noSpaces = T,
                        test = T,
                        contDigits = 1,
                        printToggle = F,
                        dropEqual = T,
                        explain = T)

# Convert to dataframe
tab_hp_closest_df <- as.data.frame(tab_plot_hp_closest) %>% rownames_to_column(var = "Variable")

# Set Table header
header <- str_squish(str_remove("Table 1. Participant demographics at the time the stool sample closest to AD biomarker assessment date was collected, stratified by AD status", "\n"))
# Set Table footer
footer <- str_squish(str_remove("Numbers are either No. or mean values", "\n"))
# Set custom_tab() defaults
customtab_defaults()
# Create the flextable object
flextable_processed_samples <- custom_tab(tab_hp_closest_df, header, footer)

# Save as word .docx
save_as_docx(flextable_processed_samples, path = "Table1.docx",
pr_section = prop_section(page_size = page_size(orient = "landscape"), type = "continuous"))
```


Figure 1
```{r}
# PCoA and permanova using single timepoint samples from healthy and preclinical samples closest to biomarker assessment date 
mtg_single_tp_hp_meta <- data.frame(phyloseq::sample_data(ps_mtg_single_tp))
ps.ordination <- phyloseq::ordinate(ps_mtg_single_tp, method='PCoA', distance='unifrac')

coordinates <- data.frame(ps.ordination$vectors)

df <- data.frame(Coord1 = coordinates$Axis.1,
                 Coord2 = coordinates$Axis.2,
                 Age = mtg_single_tp_hp_meta[, "Age"],
                 Sex = mtg_single_tp_hp_meta[, "Sex"],
                 BMI = mtg_single_tp_hp_meta[, "BMI"],
                 Copies_of_APOE4_allele = mtg_single_tp_hp_meta[, "apoe4_carriage"],
                 Diabetes = mtg_single_tp_hp_meta[, "Diabetes"],
                 Hypertension = mtg_single_tp_hp_meta[, "Hypertension"],
                 Days_since_biomarker_assessment = mtg_single_tp_hp_meta[, "Days_since_biomarker_assessment"],
                 Group = mtg_single_tp_hp_meta[, "AD_Status"])
df <- df %>%
  filter(!is.na(Copies_of_APOE4_allele))

axis1.model <- lm(Coord1 ~ Age + Sex + BMI + Copies_of_APOE4_allele + Diabetes + Hypertension + Group, data=df)
axis1_summary <- tidy(axis1.model)

# Type III ANOVA to assess the effect of AD_Status on PCoA1 scores while adjusting for covariates
## The effect of AD_Status was marginally significant (F=4.11, p=0.044), indicating that amyloid positivity explained a significant portion of the variation in taxonomic composition along PCoA axis 1. 
axis1_result <- car::Anova(axis1.model, type = 3) 
write.csv(axis1_result, "260406_PCoA1_hp_typeIII_ANOVA_results.csv", row.names = T)

axis2.model <- lm(Coord2 ~ Age + Sex + BMI + Copies_of_APOE4_allele + Diabetes + Hypertension + Group, data=df)
summary(axis2.model)
axis2_summary <- tidy(axis2.model)
write.csv(axis2_summary, "260306_PCoA2_hp_LM_results.csv", row.names = T)

## The effect of AD_Status was trending towards significance (F=3.27, p=0.072), indicating that amyloid positivity explained a significant portion of the variation in taxonomic composition along PCoA axis 2.
axis2_result <- car::Anova(axis2.model, type = 3)
write.csv(axis2_result, "260306_PCoA2_hp_typeIII_ANOVA_results.csv", row.names = T)


ordination.plot <- phyloseq::plot_ordination(ps_mtg_single_tp, ps.ordination,
                                             type="samples", color="AD_Status") +
  geom_point(size=3, shape=19) +
  stat_ellipse() +
  scale_color_manual(values=custom_colors_hp, name = "AD Status") + 
  labs(x = "PCoA1 (21.8%)", y = "PCoA2 (10%)") +
  theme_classic() +
  theme(
    axis.text.y = element_text(size=14),
    axis.text.x = element_text(size=14),
    axis.title.x = element_text(size=14),
    axis.title.y = element_text(size=14),
    legend.title = element_text(size=14),
    legend.text = element_text(size=14),
    legend.position = "left",
    plot.title = element_text(size=15),
    panel.border = element_rect(fill = NA, colour = "black", size = 3)
  )
# Add ggmarginal 
ordination.plot <- ggMarginal(ordination.plot, type = 'boxplot', groupFill = TRUE, size = 10)
ggsave("260306_Fig1A.pdf", plot = ordination.plot, bg = "transparent", width = 9, height = 6)

# Add the Unifrac PCoA 1 and 2 axes to the phyloseq object metadata
sample_data(ps_mtg_single_tp)$Tax.PCoA1 <- ps.ordination$vectors[,1]
sample_data(ps_mtg_single_tp)$Tax.PCoA2 <- ps.ordination$vectors[,2]

# PERMANOVA with all significantly different variables based on Table 1
## N = 197 (drops 1 sample without APOE data)
perm_mtg_single_tp_hp <- permanova_wrapper(ps_mtg_single_tp, values=c("Healthy","Preclinical"), var = "AD_Status", adonis_values = c("Age","Sex","BMI","apoe4_carriage","Diabetes","Hypertension","AD_Status"), ss_method = "margin") #p-value = 0.028
write.csv(perm_mtg_single_tp_hp@permanova, "260306_PCoA_permanova.csv")

######### Run MaAslin2 and create Fig. 1C ############
ps_mtg_single_tp_filt <- ps_mtg_single_tp %>% subset_samples(!is.na(apoe4_carriage))

maaslin_wrapper(ps_mtg_single_tp_filt, random_effects_vector=c(), reference_vector = c("AD_Status,Healthy", "apoe4_carriage,0", "Sex,M", "Diabetes,0", "Hypertension,0"), fixed_effects_vector = c("Age","Sex","BMI","apoe4_carriage","Diabetes","Hypertension","AD_Status"), anal_method = "NEGBIN", core = 8)

num_samples <- length(ps_mtg_single_tp_filt@sam_data$Sample_ID)
sample_cutoff <- num_samples*.10
maaslin_output <- "~/Library/CloudStorage/Box-Box/2026_JianMark_ADGutOmics2/Figures/260306_figures/Figure1/MaAslin2_results/significant_results.tsv"

maaslin_df <- read_tsv(maaslin_output) %>% filter(metadata=="AD_Status") %>% mutate(abs_coef=abs(coef)) %>% mutate(qval = p.adjust(pval, method = 'BH')) %>% filter(N.not.0>sample_cutoff) %>% mutate(feature=gsub(".*s__", "", feature)) %>% mutate(value=as.factor(value)) 

median <- median(maaslin_df$abs_coef)
maaslin_plot_subset <- subset(maaslin_df, N.not.0 > sample_cutoff & abs(coef) > median)
maaslin_plot_subset <- maaslin_plot_subset %>% arrange(desc(coef))
#obtain min and max coefficient values (needed for plotting later)
min <- min(maaslin_plot_subset$coef)
max <- max(maaslin_plot_subset$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
maaslin_plot_subset <- maaslin_plot_subset %>% mutate(species = fct_reorder(feature, coef))

maaslin_plot_subset$coefmax <- maaslin_plot_subset$coef + maaslin_plot_subset$stderr
maaslin_plot_subset$coefmin <- maaslin_plot_subset$coef - maaslin_plot_subset$stderr

maaslin_plot_subset$CI95upper <- maaslin_plot_subset$coef + 1.96*maaslin_plot_subset$stderr
maaslin_plot_subset$CI95lower <- maaslin_plot_subset$coef - 1.96*maaslin_plot_subset$stderr
write.csv(maaslin_plot_subset, "260306_Fig1C_coeff_values.csv", row.names = F)

# Plot coefficients
x_min_lim <- (floor(min / 5) * 6) 
x_max_lim <- (ceiling(max / 5) * 6) 
colors_list <- c("#eebe61","#49b9a5")

coeff_p <- ggplot(maaslin_plot_subset, aes(x=coef, y=species, color=as.factor(coef<0)))+
  geom_point(size = 3)+
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  scale_color_manual(values=colors_list)+
  theme_linedraw()+
  theme(axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=18),
        axis.title.x = element_text(size=17),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.position = "bottom",
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", linewidth=1))+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  xlab('Coefficient')
ggsave("260306_Fig1C_coeff.pdf", plot = coeff_p, bg = "transparent", width = 11, height = 12)

#### Generate prevalence plot ###
### Access species abundance data 
abund <- psmelt(ps_mtg_single_tp_filt)

abund <- abund %>% mutate(OTU=gsub(".*s__", "", OTU))
diff_feat_abund <- subset(abund, OTU %in% maaslin_plot_subset$feature)

# Re-order feature abs abundances according to coefficient value
# This order is preserved in filename, which was previously ordered: 
order <- data.frame(OTU = unique(maaslin_plot_subset$feature))
# sort = F makes sure that the order is based on order_index df which is ordered by coefficient value
abund.df <- merge(order, diff_feat_abund, by="OTU", sort = F)
abund.df$OTU <- as.factor(abund.df$OTU)
abund.df <- abund.df %>% mutate(OTU = fct_relevel(OTU, maaslin_plot_subset$feature))

# For each significant feature in each sample, set AbunNotZero == 1 if its 
# abundance is greater than 0. 
abund.df$AbunNotZero <- 1*(abund.df$Abundance > 0)

# Summarize pathway prevalence by variable of interest
abund.df.n0 <- abund.df %>% 
  group_by(OTU, AD_Status) %>%
  summarise(N = n(), SumNotZero = sum(AbunNotZero)) %>%
  mutate(PercentNotZero = 100*(SumNotZero / N)) %>%
  ungroup()
write.csv(abund.df.n0, "260306_maaslin2_species_prevalence.csv", 
          row.names = F)

prev_p <- ggplot(abund.df.n0, aes(x = AD_Status, y=OTU, fill=PercentNotZero))+
  geom_tile()+
  scale_fill_distiller(palette ='RdBu', type='seq', breaks = c(0,50,100), 'Prevalence (%)')+
  theme_classic()+
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(size=18, angle=45, hjust=1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=17),
        legend.text = element_text(size = 18),
        panel.border = element_rect(fill=NA, colour="black", size=1))+
  scale_y_discrete(limits=rev)
ggsave("260306_Fig1C_prev.pdf", plot = prev_p, bg = "transparent", width = 11, height = 12)

combined_maaslin_p <- coeff_p | prev_p 
ggsave("260306_Fig1B_MaAslin2_taxa_plot.pdf", plot = combined_maaslin_p, bg = "transparent", width = 16, height = 10)


###### Comparing output with Ferreiro et al. 2023 ########
af_results <- read_excel("~/Library/CloudStorage/Box-Box/DantasLab/AD_microbiome/Aura/scitranslmed.abo2984_data_files_s1_to_s4/Aura_MaAslin2_species_Amyloid_positive_results.xlsx")

af_results2 <- af_results %>% select(feature, coef) %>% dplyr::rename(af_coef = coef) %>% mutate(feature=gsub(".*s__", "", feature))

common_results <- merge(maaslin_df, af_results2, by = "feature", all.x = T) %>% arrange(desc(coef)) %>%
  mutate(consistent = case_when(is.na(coef) | is.na(af_coef) ~ NA_character_,
                                coef > 0 & af_coef > 0 ~ "true_pos",
                                coef < 0 & af_coef < 0 ~ "true_neg",
                                TRUE ~ "false"))
consistent_prop <- common_results %>%  count(consistent) %>% mutate(proportion = n / sum(n))

maaslin_df <- maaslin_df %>% arrange(desc(coef))

####### Prepare supplementary table 4 #######
maaslin_df_all <- read_tsv(maaslin_output) %>% mutate(abs_coef=abs(coef)) %>% mutate(qval = p.adjust(pval, method = 'BH')) %>% filter(N.not.0>sample_cutoff) %>% mutate(feature=gsub(".*s__", "", feature)) %>% mutate(value=as.factor(value)) %>% arrange(metadata, desc(coef))

## Add a column that indicates whether the taxa was identified in Ferreiro et al. 2023
maaslin_df_all <- maaslin_df_all %>% mutate(Identified_in_Ferreiro_et_al = ifelse(metadata == "AD_Status", ifelse(is.na(common_results$consistent), FALSE, TRUE), NA_character_))
write.csv(maaslin_df_all, "260306_MaAslin2_taxa_Supplementary_Table4.csv", row.names = F)

```

Extended Figure 1
```{r}
#### Compare alpha diversity between groups ###
alpha_div_mtg_single_tp <- alpha_div_wrapper_adjust_var(ps_mtg_single_tp, 
                                                        values=c("Healthy", "Preclinical"),
                                                        color_vector = custom_colors_hp,
                                                        save_adiv_plot = TRUE)
sample_data(ps_mtg_single_tp)$Tax_Richness <- alpha_div_mtg_single_tp$Richness
sample_data(ps_mtg_single_tp)$Tax_Shannon <- alpha_div_mtg_single_tp$Shannon
sample_data(ps_mtg_single_tp)$Tax_Simpson <- alpha_div_mtg_single_tp$Simpson
```


Figure 2A-C, Extended Data Figure 2A-D, Figures 4A-C, Extended Data Figure 5A-D: Network analysis
```{r}
# Prepare phyloseq objects for SECOM
ps_mtg_last_h <- ps_mtg_last_tp_hp %>% subset_samples(AD_Status == "Healthy") 
ps_mtg_last_p <- ps_mtg_last_tp_hp %>% subset_samples(AD_Status == "Preclinical")
ps_mtg_closest_h <- ps_mtg_single_tp %>% subset_samples(AD_Status == "Healthy")
ps_mtg_closest_p <- ps_mtg_single_tp %>% subset_samples(AD_Status == "Preclinical")

ps_mtg_last_p_meta <- data.frame(sample_data(ps_mtg_last_p))
ps_mtg_last_h_meta <- data.frame(sample_data(ps_mtg_last_h))
ps_mtg_closest_h_meta <- data.frame(sample_data(ps_mtg_closest_h))
ps_mtg_closest_p_meta <- data.frame(sample_data(ps_mtg_closest_p))


tso_mtg_h_last = mia::makeTreeSummarizedExperimentFromPhyloseq(ps_mtg_last_h)  # 64 samples
tso_mtg_p_last = mia::makeTreeSummarizedExperimentFromPhyloseq(ps_mtg_last_p) # 25 samples
tso_mtg_h_closest = mia::makeTreeSummarizedExperimentFromPhyloseq(ps_mtg_closest_h) # 137 samples
tso_mtg_p_closest = mia::makeTreeSummarizedExperimentFromPhyloseq(ps_mtg_closest_p) # 61 samples

# Run SECOM
set.seed(123)
secom_h_closest = secom_linear(data = list(tso_mtg_h_closest), assay_name = "counts",
                          tax_level = NULL, pseudo = 0, 
                          prv_cut = 0, lib_cut = 0, corr_cut = 0.6, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, max_p = 0.05, n_cl = 2)
secom_p_closest = secom_linear(data = list(tso_mtg_p_closest), assay_name = "counts",
                          tax_level = NULL, pseudo = 0, 
                          prv_cut = 0, lib_cut = 0, corr_cut = 0.6, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, max_p = 0.05, n_cl = 2)

secom_h_last = secom_linear(data = list(tso_mtg_h_last), assay_name = "counts",
                          tax_level = NULL, pseudo = 0, 
                          prv_cut = 0, lib_cut = 0, corr_cut = 0.6, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, max_p = 0.05, n_cl = 2)

secom_p_last = secom_linear(data = list(tso_mtg_p_last), assay_name = "counts",
                          tax_level = NULL, pseudo = 0, 
                          prv_cut = 0, lib_cut = 0, corr_cut = 0.6, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, max_p = 0.05, n_cl = 2)

#### Networks were created with the computed nodes and edges below using Gephi 
set.seed(123)
P_closest_res_linear_network <- calc_secom_network_properties(secom_p_closest, overlap_threshold = 10, cor_threshold = 0.6) 
H_closest_res_linear_network <- calc_secom_network_properties(secom_h_closest, overlap_threshold = 10, cor_threshold = 0.6)

P_last_res_linear_network <- calc_secom_network_properties(secom_p_last, overlap_threshold = 10, cor_threshold = 0.5)
H_last_res_linear_network <- calc_secom_network_properties(secom_h_last, overlap_threshold = 10, cor_threshold = 0.5) 

h_closest_node <- H_closest_res_linear_network$nodes
h_closest_node <- h_closest_node %>% 
  mutate(id=gsub(".*s__", "", id),
         Label=gsub(".*s__", "", Label))


p_closest_node <- P_closest_res_linear_network$nodes
p_closest_node <- p_closest_node %>% 
  mutate(id=gsub(".*s__", "", id),
         Label=gsub(".*s__", "", Label))
#write.csv(h_closest_node, "260306_preclinical_closest_nodes.csv", row.names = F)

h_last_node <- H_last_res_linear_network$nodes
h_last_node <- h_last_node %>% 
  mutate(id=gsub(".*s__", "", id),
         Label=gsub(".*s__", "", Label))
#write.csv(h_last_node, "260306_healthy_last_nodes.csv", row.names = F)

p_last_node <- P_last_res_linear_network$nodes
p_last_node <- p_last_node %>% 
  mutate(id=gsub(".*s__", "", id),
         Label=gsub(".*s__", "", Label))
#write.csv(p_last_node, "260306_preclinical_last_nodes.csv", row.names = F)

h_closest_edges <- H_closest_res_linear_network$edges
h_closest_edges <- h_closest_edges %>% 
  mutate(Source=gsub(".*s__", "", Source),
         Target=gsub(".*s__", "", Target))
#write.csv(h_closest_edges, "260306_healthy_closest_edges.csv", row.names = F)

p_closest_edges <- P_closest_res_linear_network$edges
p_closest_edges <- p_closest_edges %>% 
  mutate(Source=gsub(".*s__", "", Source),
         Target=gsub(".*s__", "", Target))
#write.csv(p_closest_edges, "260306_preclincial_closest_edges.csv", row.names = F)

h_last_edges <- H_last_res_linear_network$edges
h_last_edges <- h_last_edges %>% 
  mutate(Source=gsub(".*s__", "", Source),
         Target=gsub(".*s__", "", Target))
#write.csv(h_last_edges, "260306_healthy_last_edges.csv", row.names = F)

p_last_edges <- P_last_res_linear_network$edges
p_last_edges <- p_last_edges %>% 
  mutate(Source=gsub(".*s__", "", Source),
         Target=gsub(".*s__", "", Target))
#write.csv(p_last_edges, "260306_preclinical_last_edges.csv", row.names = F)

##############----------- Create abbreviated name column for display in Gephi --------------###
h_closest_node <- h_closest_node %>%
  rowwise() %>%
  mutate(
    parts = list(str_split(id, "_")[[1]]),  # wrap in list for rowwise compatibility
    display_label = case_when(
      length(parts) == 2 ~ paste0(substr(parts[1], 1, 1), ". ", parts[2]),
      length(parts) > 2 & "sp" %in% parts ~ paste0(parts[1], " sp. ", paste(parts[3:length(parts)], collapse = ":")),
      length(parts) > 2 & !"sp" %in% parts ~ paste(paste(parts[1:2], collapse = " "), paste(parts[3:length(parts)], collapse = ":"), sep = " "),
      TRUE ~ id)) %>% ungroup() %>% select(-parts)
write.csv(h_closest_node, "260306_healthy_closest_nodes.csv", row.names = F) 

p_closest_node <- p_closest_node %>%
  rowwise() %>%
  mutate(
    parts = list(str_split(id, "_")[[1]]),  # wrap in list for rowwise compatibility
    display_label = case_when(
      length(parts) == 2 ~ paste0(substr(parts[1], 1, 1), ". ", parts[2]),
      length(parts) > 2 & "sp" %in% parts ~ paste0(parts[1], " sp. ", paste(parts[3:length(parts)], collapse = ":")),
      length(parts) > 2 & !"sp" %in% parts ~ paste(paste(parts[1:2], collapse = " "), paste(parts[3:length(parts)], collapse = ":"), sep = " "),TRUE ~ id)) %>% ungroup() %>% select(-parts)
write.csv(p_closest_node, "260306_preclinical_closest_nodes.csv", row.names = F)

h_last_node <- h_last_node %>%
  rowwise() %>%
  mutate(
    parts = list(str_split(id, "_")[[1]]),  # wrap in list for rowwise compatibility
    display_label = case_when(
      length(parts) == 2 ~ paste0(substr(parts[1], 1, 1), ". ", parts[2]),
      length(parts) > 2 & "sp" %in% parts ~ paste0(parts[1], " sp. ", paste(parts[3:length(parts)], collapse = ":")),
      length(parts) > 2 & !"sp" %in% parts ~ paste(paste(parts[1:2], collapse = " "), paste(parts[3:length(parts)], collapse = ":"), sep = " "),TRUE ~ id)) %>% ungroup() %>% select(-parts)
write.csv(h_last_node, "260306_healthy_last_nodes.csv", row.names = F)

p_last_node <- p_last_node %>%
  rowwise() %>%
  mutate(
    parts = list(str_split(id, "_")[[1]]),  # wrap in list for rowwise compatibility
    display_label = case_when(
      length(parts) == 2 ~ paste0(substr(parts[1], 1, 1), ". ", parts[2]),
      length(parts) > 2 & "sp" %in% parts ~ paste0(parts[1], " sp. ", paste(parts[3:length(parts)], collapse = ":")),
      length(parts) > 2 & !"sp" %in% parts ~ paste(paste(parts[1:2], collapse = " "), paste(parts[3:length(parts)], collapse = ":"), sep = " "),TRUE ~ id)) %>% ungroup() %>% select(-parts)
write.csv(p_last_node, "260306_preclinical_last_nodes.csv", row.names = F)

####### Compare hubbiness metrics between networks: Cross-sectional (Fig D)
hub_metrics_h_closest <- h_closest_node %>% mutate(Group = "Healthy")
hub_metrics_p_closest <- p_closest_node %>% mutate(Group = "Preclinical")
hub_metrics_closest <- rbind(hub_metrics_h_closest, hub_metrics_p_closest)
hub_metrics_closest$Group <- as.factor(hub_metrics_closest$Group)
hub_metrics_closest$Group <- factor(hub_metrics_closest$Group, levels=c("Healthy","Preclinical"))

bp_closest_degree <- ggplot(hub_metrics_closest, aes(x = Group, y = degree, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Degree") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy", "Preclinical")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Fig2C_degree.pdf", plot = bp_closest_degree, bg = "transparent", width = 10, height = 8)


bp_closest_betweenness <- ggplot(hub_metrics_closest, aes(x = Group, y = betweenness, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Betweenness Centrality",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy", "Preclinical")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 5, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Extended_Fig_2A.pdf", plot = bp_closest_betweenness, bg = "transparent", width = 10, height = 8)

bp_closest_closeness <- ggplot(hub_metrics_closest, aes(x = Group, y = closeness, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Closeness Centrality",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")
        ) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy", "Preclinical")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Fig2C_closeness.pdf", plot = bp_closest_closeness, bg = "transparent", width = 10, height = 8)

bp_closest_coreness <- ggplot(hub_metrics_closest, aes(x = Group, y = Coreness, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Coreness",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")
        ) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy", "Preclinical")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Figure2C_coreness.pdf", plot = bp_closest_coreness, bg = "transparent", width = 10, height = 8)

bp_closest_efficiency <- ggplot(hub_metrics_closest, aes(x = Group, y = local_efficiency, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Local Efficiency",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")
        ) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy", "Preclinical")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Figure2C_local_effiency.pdf", plot = bp_closest_efficiency, bg = "transparent", width = 10, height = 8)

```

Fig 2D: Comparing guild abundance at the closest time point
```{r}
ps_mtg_single_tp_meta <- data.frame(phyloseq::sample_data(ps_mtg_single_tp))

# Add a tiny number to zero values since plot below log transforms guild abundance values
ps_mtg_single_tp_meta <- ps_mtg_single_tp_meta %>% mutate(guild_abundance_adj = guild_abundance + 1e-6)

## Plot Figure 2D
closest_hp_abundance_plot <- ggplot(ps_mtg_single_tp_meta, aes(x = AD_Status, y = guild_abundance_adj, color = AD_Status)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.8, size = 1) +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  scale_color_manual(values = custom_colors_hp) +
  theme_bw() +
  theme(
    text = element_text(size = 16),       
    axis.title = element_text(size = 18), 
    axis.text = element_text(size = 14),  
    legend.position = "none") +
  scale_y_log10() +
  labs(
    y = "Guild Abundance",
    x = "AD Status")
ggsave("260402_MTG_guild_abund_comparison_hp.pdf", plot = closest_hp_abundance_plot,bg = "transparent", width = 8, height = 6)

```


Figure 3: Stool Inflammatory markers
```{r}
ps_mtg_single_tp_meta$apoe4_carriage <- relevel(factor(ps_mtg_single_tp_meta$apoe4_carriage), ref = "0")
ps_mtg_single_tp_meta$Diabetes <- relevel(factor(ps_mtg_single_tp_meta$Diabetes), ref = "0")
ps_mtg_single_tp_meta$Hypertension <- relevel(factor(ps_mtg_single_tp_meta$Hypertension), ref = "0")
ps_mtg_single_tp_meta$AD_Status <- relevel(factor(ps_mtg_single_tp_meta$AD_Status), ref = "Healthy")

df_single_tp_hp_ngal <- ps_mtg_single_tp_meta %>%
  select(c(Sample_ID, ID, log2_NGAL, Age, Sex, BMI, apoe4_carriage, Diabetes, Hypertension, AD_Status, combined.centiloid)) %>%
  filter(complete.cases(.)) # Eliminating rows with NAs 

NGAL_sample_sizes_ad_status <- df_single_tp_hp_ngal %>%
  filter(!is.na(log2_NGAL)) %>%
  group_by(AD_Status) %>%
  summarise(n = n(), .groups = 'drop')  # Create a summary with sample sizes

options(na.action='na.fail')
ngal.model <- lm(log2_NGAL ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + AD_Status, data = df_single_tp_hp_ngal)
summary(ngal.model)
ngal_model_summary <- tidy(ngal.model)
#ngal_model_summary$q.value <- p.adjust(ngal_model_summary$p.value, method = "BH")
write.csv(ngal_model_summary, "260309_log2_NGAL_singletp_lm_summary_hp.csv", row.names = F)

#### Adjusted EMMs
ngal_emm <- pairs(emmeans(ngal.model, ~ AD_Status), adjust = "BH")
summary(ngal_emm)

ngal_ad_comp <- as.data.frame(ngal_emm)
#write.csv(ngal_ad_comp, "260309_NGAL_AD_LMM_emmeans_results.csv", row.names = F)
ngal_ad_sig <- ngal_ad_comp %>%
  filter(p.value < 0.05) %>%
  mutate(
    label = paste0("q = ", signif(p.value, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast)))

### LMM BH-corrected p-values
p_NGAL_hp <- ggplot(df_single_tp_hp_ngal, aes(x = AD_Status, y = log2_NGAL, fill = AD_Status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = AD_Status),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Log2 Stool NGAL (ug/mL)",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.title = element_text(size = 16),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")   # Prevents clipping of points at the bottom

y_positions <- df_single_tp_hp_ngal %>%
  group_by(AD_Status) %>%
  summarise(y_max = max(log2_NGAL, na.rm = TRUE)) %>%
  deframe()

step_height <- 0.5
for(i in seq_len(nrow(ngal_ad_sig))) {
  g1 <- ngal_ad_sig$group1[i]
  g2 <- ngal_ad_sig$group2[i]
  label <- ngal_ad_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(df_single_tp_hp_ngal$AD_Status) == g1)
  x2 <- which(levels(df_single_tp_hp_ngal$AD_Status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_NGAL_hp <- p_NGAL_hp + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.1, label = label, size = 4.5)
}
ggsave("260406_NGAL_LMM_pvalues_AD_Status_HP.pdf", plot = p_NGAL_hp, bg = "transparent", width = 10, height = 8) 

############# Calprotectin ###########
single_tp_hp_calp <- ps_mtg_single_tp_meta %>%
  select(c(Sample_ID, ID, log2_Calp, Age, Sex, BMI, apoe4_carriage, Diabetes, Hypertension, AD_Status, combined.centiloid)) %>%
  filter(complete.cases(.)) # Eliminating rows with NAs 

options(na.action='na.fail')
calp.model <- lm(log2_Calp ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + AD_Status, data = single_tp_hp_calp)
summary(calp.model)

calp_model_summary <- tidy(calp.model)
#calp_model_summary$q.value <- p.adjust(calp_model_summary$p.value, method = "BH")
write.csv(calp_model_summary, "260406_log2_Calprotectin_singletp_lm_summary.csv", row.names = F)

#### Adjusted EMMs
calp_emm <- pairs(emmeans(calp.model, ~ AD_Status), adjust = "none")
summary(calp_emm)
calp_ad_comp <- as.data.frame(calp_emm)
write.csv(calp_ad_comp, "260406_Calp_AD_LMM_emmeans_results.csv", row.names = F)

calp_ad_sig <- calp_ad_comp %>%
  filter(p.value < 0.05) %>%
  mutate(
    label = paste0("q = ", signif(p.value, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast)))

p_Calp_hp <- ggplot(single_tp_hp_calp, aes(x = AD_Status, y = log2_Calp, fill = AD_Status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = AD_Status),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Log2 Stool Calprotectin (ug/mL)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.title = element_text(size = 16),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off") 

y_positions <- single_tp_hp_calp %>%
  group_by(AD_Status) %>%
  summarise(y_max = max(log2_Calp, na.rm = TRUE)) %>%
  deframe()

step_height <- 0.5
for(i in seq_len(nrow(calp_ad_sig))) {
  g1 <- calp_ad_sig$group1[i]
  g2 <- calp_ad_sig$group2[i]
  label <- calp_ad_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(single_tp_hp_calp$AD_Status) == g1)
  x2 <- which(levels(single_tp_hp_calp$AD_Status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_Calp_hp <- p_Calp_hp + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.25, label = label, size = 4.5)
}
ggsave("260406_Calp_LMM_pvalues_AD_Status_HP.pdf", plot = p_Calp_hp, bg = "transparent", width = 10, height = 8) 

### Combined NGAL + Calprotectin plot
combined_inflamm_plot_hor <- p_NGAL_hp | p_Calp_hp
ggsave("Fig_3A.pdf", plot = combined_inflamm_plot_hor, bg = "transparent", width = 14, height = 7)

############## By AT Status ##############
single_tp_at_ngal <- ps_mtg_single_tp_meta %>%
  select(c(Sample_ID, ID, log2_NGAL, Age, Sex, BMI, apoe4_carriage, Diabetes, Hypertension, amyloid_tau_status)) %>%
  filter(complete.cases(.)) # Eliminating rows with NAs 

options(na.action='na.fail')
ngal_at.model <- lm(log2_NGAL ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + amyloid_tau_status, data = single_tp_at_ngal)
summary(ngal_at.model)
ngal_at_model_summary <- tidy(ngal_at.model)
ngal_at_model_summary$q.value <- p.adjust(ngal_at_model_summary$p.value, method = "BH")
#write.csv(ngal_at_model_summary, "260309_log2_NGAL_singletp_lm_summary_ALL.csv", row.names = F)

## Adjusted means (EMMs)
ngal_at_emm <- pairs(emmeans(ngal_at.model, ~ amyloid_tau_status), adjust = "none")
summary(ngal_at_emm)
ngal_at_comp <- as.data.frame(ngal_at_emm)
ngal_at_comp$BH.adj.p <- p.adjust(ngal_at_comp$p.value, method = "BH")
write.csv(ngal_at_comp, "260406_NGAL_AT_LMM_emmeans_results.csv", row.names = F)
ngal_at_sig <- ngal_at_comp %>%
  filter(p.value < 0.05) %>%
  mutate(
    label = paste0("q = ", signif(BH.adj.p, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast)))
#write.csv(ngal_at_sig, "260309_NGAL_AT_BH_adj_LMM_emmeans_signif_results.csv", row.names = F)

p_NGAL_at_status <- ggplot(single_tp_at_ngal, aes(x = amyloid_tau_status, y = log2_NGAL, fill = amyloid_tau_status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = amyloid_tau_status),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_at_status) + 
  scale_color_manual(values = custom_colors_at_status) +
  labs(x = "Amyloid Tau Status",
       y = "Log2 Stool NGAL (ug/mL)",
       fill = "Amyloid Tau Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 13),
        legend.position = "none",
        legend.background = element_rect(fill = "white", color = "black"),
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")   # Prevents clipping of points at the bottom

y_positions <- single_tp_at_ngal %>%
  group_by(amyloid_tau_status) %>%
  summarise(y_max = max(log2_NGAL, na.rm = TRUE)) %>%
  deframe()

step_height <- 0.5
for(i in seq_len(nrow(ngal_at_sig))) {
  g1 <- ngal_at_sig$group1[i]
  g2 <- ngal_at_sig$group2[i]
  label <- ngal_at_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(single_tp_at_ngal$amyloid_tau_status) == g1)
  x2 <- which(levels(single_tp_at_ngal$amyloid_tau_status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_NGAL_at_status <- p_NGAL_at_status + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.2, label = label, size = 4.5)
}
#ggsave("260309_NGAL_LMM_pvalues_AT_Status_HP.pdf", plot = p_NGAL_at_status, bg = "transparent", width = 10, height = 8) 

############# Calprotectin ##########
single_tp_at_calp <- ps_mtg_single_tp_meta %>%
  select(c(Sample_ID, ID, log2_Calp, Age, Sex, BMI, apoe4_carriage, Diabetes, Hypertension, amyloid_tau_status)) %>%
  filter(complete.cases(.)) # Eliminating rows with NAs 

options(na.action='na.fail')
calp_at.model <- lm(log2_Calp ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + amyloid_tau_status, data = single_tp_at_calp)
summary(calp_at.model)
calp_at_model_summary <- tidy(calp_at.model)
calp_at_model_summary$q.value <- p.adjust(calp_at_model_summary$p.value, method = "BH")
write.csv(calp_at_model_summary, "260406_log2_Calp_singletp_lm_summary_HP.csv", row.names = F)

## Adjusted means (EMMs)
calp_at_emm <- pairs(emmeans(calp_at.model, ~ amyloid_tau_status), adjust = "none")
summary(calp_at_emm)
calp_at_comp <- as.data.frame(calp_at_emm)
calp_at_comp$BH.adj.p <- p.adjust(calp_at_comp$p.value, method = "BH")
write.csv(calp_at_comp, "260406_Calp_AT_LMM_emmeans_results.csv", row.names = F)
calp_at_sig <- calp_at_comp %>%
  filter(p.value < 0.05) %>%
  mutate(
    label = paste0("q = ", signif(BH.adj.p, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast))
  )
#write.csv(calp_at_sig, "260309_Calp_AT_BH_adj_LMM_emmeans_signif_results.csv", row.names = F)

p_Calp_at_status <- ggplot(single_tp_at_calp, aes(x = amyloid_tau_status, y = log2_Calp, fill = amyloid_tau_status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = single_tp_at_calp),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_at_status) + 
  scale_color_manual(values = custom_colors_at_status) +
  labs(x = "Amyloid Tau Status",
       y = "Log2 Stool Calprotectin (ug/mL)",
       fill = "Amyloid Tau Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 13),
        legend.position = "none",
        legend.background = element_rect(fill = "white", color = "black"),
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")   # Prevents clipping of points at the bottom

y_positions <- ps_mtg_single_tp_meta %>%
  group_by(amyloid_tau_status) %>%
  summarise(y_max = max(log2_Calp, na.rm = TRUE)) %>%
  deframe()

step_height <- 1.7
for(i in seq_len(nrow(calp_at_sig))) {
  g1 <- calp_at_sig$group1[i]
  g2 <- calp_at_sig$group2[i]
  label <- calp_at_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(single_tp_at_calp$amyloid_tau_status) == g1)
  x2 <- which(levels(single_tp_at_calp$amyloid_tau_status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_Calp_at_status <- p_Calp_at_status + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.8, label = label, size = 4.5)
}
#ggsave("260309_Calp_LMM_pvalues_AT_Status_HP.pdf", plot = p_Calp_at_status, bg = "transparent", width = 10, height = 8) 

### Combined NGAL + Calprotectin plot
combined_inflamm_plot_hor_AT <- p_NGAL_at_status | p_Calp_at_status
ggsave("ExtendedFig_3A_hor_hp.pdf", plot = combined_inflamm_plot_hor_AT, bg = "transparent", width = 14, height = 8)

###############----------- Extended Data Figure 4B and C --------------#############
#### Correlations between inflammation markers and PET amyloid-beta ####
# Calculate correlation
cor_test <- cor.test(df_single_tp_hp_ngal$combined.centiloid, df_single_tp_hp_ngal$log2_NGAL)
r_val <- round(cor_test$estimate, 2)
p_val <- round(cor_test$p.value, 5)

# Plot
ngal_cor <- ggplot(df_single_tp_hp_ngal, aes(x = combined.centiloid, y = log2_NGAL)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", color = "steelblue", fill = "grey80") +
  annotate("text", x = Inf, y = Inf, 
           label = paste0("R = ", r_val, ", p = ", p_val),
           hjust = 1.1, vjust = 1.5, size = 5) +
  labs(x = "Centiloid",
       y = "Log2 NGAL (ug/mL)") +
  theme_minimal() %>%
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_text(size = 12))
ggsave("260309_NGAL_Centiloid_cor.pdf", plot = ngal_cor, bg = "transparent", width = 8, height = 6) 

cor_test <- cor.test(single_tp_hp_calp$combined.centiloid, single_tp_hp_calp$log2_Calp)
r_val <- round(cor_test$estimate, 2)
p_val <- round(cor_test$p.value, 5)

# Plot
calp_cor <- ggplot(single_tp_hp_calp, aes(x = combined.centiloid, y = log2_Calp)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", color = "steelblue", fill = "grey80") +
  annotate("text", x = Inf, y = Inf, 
           label = paste0("R = ", r_val, ", p = ", p_val),
           hjust = 1.1, vjust = 1.5, size = 5) +
  labs(x = "Centiloid",
       y = "Log2 Calprotectin (ug/mL)") +
  theme_minimal() %>%
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_text(size = 12))
ggsave("260309_Calprotectin_Centiloid_cor.pdf", plot = calp_cor, bg = "transparent", width = 8, height = 6) 

#### Longitudinal stool inflammation markers (Need to move this out of this script) #####
ps_mtg_long_meta <- data.frame(sample_data(ps_MTG))

elisa_long_df <- ps_mtg_long_meta %>%
  select(c(Sample_ID, ID, Time_diff_from_baseline_mo, log2_Calp, log2_NGAL, Age, Sex, BMI, apoe4_carriage, Diabetes, Hypertension, AD_Status, amyloid_tau_status)) %>%
  filter(complete.cases(.)) # Eliminating rows with NAs 

elisa_long_df$apoe4_carriage <- relevel(factor(elisa_long_df$apoe4_carriage), ref = "0")
elisa_long_df$Diabetes <- relevel(factor(elisa_long_df$Diabetes), ref = "0")
elisa_long_df$Hypertension <- relevel(factor(elisa_long_df$Hypertension), ref = "0")
elisa_long_df$AD_Status <- relevel(factor(elisa_long_df$AD_Status), ref = "Healthy")

NGAL_long_sample_sizes_ad_status <- elisa_long_df %>%
  filter(!is.na(log2_NGAL)) %>%
  group_by(AD_Status) %>%
  summarise(n = n(), .groups = 'drop')  # Create a summary with sample sizes

library(broom.mixed)
options(na.action='na.fail')
long_ngal.model <- lmer(log2_NGAL ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + AD_Status + (1 | ID), data = elisa_long_df, REML = FALSE)
summary(long_ngal.model)
long_ngal_lmm <- tidy(long_ngal.model)
write.csv(long_ngal_lmm, "260309_log2_Calp_long_lm_summary_HP.csv", row.names = F)

#### Adjusted EMMs
long_ngal_emm <- pairs(emmeans(long_ngal.model, ~ AD_Status), adjust = "BH")
summary(long_ngal_emm)

long_ngal_ad_comp <- as.data.frame(long_ngal_emm)
write.csv(long_ngal_ad_comp, "260309_Long_NGAL_AD_LMM_emmeans_results.csv", row.names = F)
long_ngal_ad_sig <- long_ngal_ad_comp %>%
  mutate(
    label = paste0("q = ", signif(p.value, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast))
  )

### LMM BH-corrected p-values
p_long_NGAL_hp <- ggplot(elisa_long_df, aes(x = AD_Status, y = log2_NGAL, fill = AD_Status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = AD_Status),shape = 24, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Log2 Stool NGAL (ug/mL)",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.title = element_text(size = 16),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")    # Prevents clipping of points at the bottom

## Return here 
y_positions <- elisa_long_df %>%
  group_by(AD_Status) %>%
  summarise(y_max = max(log2_NGAL, na.rm = TRUE)) %>%
  deframe()

step_height <- 0.5
for(i in seq_len(nrow(long_ngal_ad_sig))) {
  g1 <- long_ngal_ad_sig$group1[i]
  g2 <- long_ngal_ad_sig$group2[i]
  label <- long_ngal_ad_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(elisa_long_df$AD_Status) == g1)
  x2 <- which(levels(elisa_long_df$AD_Status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_long_NGAL_hp <- p_long_NGAL_hp + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.1, label = label, size = 4.5)
}
ggsave("260309_Long_NGAL_LMM_pvalues_AD_Status_HP.pdf", plot = p_long_NGAL_hp, bg = "transparent", width = 10, height = 8) 

######## ATN Split ########
p_long_NGAL_atn <- ggplot(elisa_long_df, aes(x = amyloid_tau_status, y = log2_NGAL, fill = amyloid_tau_status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = amyloid_tau_status),shape = 24, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_at_status) +  
  scale_color_manual(values = custom_colors_at_status) +
  labs(x = "Amyloid Tau Status",
       y = "Log2 Stool NGAL (ug/mL)",
       fill = "Amyloid Tau Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.title = element_text(size = 16),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")    # Prevents clipping of points at the bottom

options(na.action='na.fail')
long_ngal_atn.model <- lmer(log2_NGAL ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + amyloid_tau_status + (1 | ID), data = elisa_long_df, REML = FALSE)
summary(long_ngal_atn.model)

#### Adjusted EMMs
long_ngal_atn_emm <- pairs(emmeans(long_ngal_atn.model, ~ amyloid_tau_status), adjust = "BH")
summary(long_ngal_atn_emm)

long_ngal_at_comp <- as.data.frame(long_ngal_atn_emm)
write.csv(long_ngal_ad_comp, "260309_long_NGAL_AT_LMM_emmeans_results.csv", row.names = F)
long_ngal_at_sig <- long_ngal_at_comp %>%
  filter(p.value < 0.25) %>%
  mutate(
    label = paste0("q = ", signif(p.value, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast)))

## Return here 
y_positions <- elisa_long_df %>%
  group_by(amyloid_tau_status) %>%
  summarise(y_max = max(log2_NGAL, na.rm = TRUE)) %>%
  deframe()

step_height <- 0.6
for(i in seq_len(nrow(long_ngal_at_sig))) {
  g1 <- long_ngal_at_sig$group1[i]
  g2 <- long_ngal_at_sig$group2[i]
  label <- long_ngal_at_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(elisa_long_df$amyloid_tau_status) == g1)
  x2 <- which(levels(elisa_long_df$amyloid_tau_status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_long_NGAL_atn <- p_long_NGAL_atn + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.2, label = label, size = 4.5)}
ggsave("260309_Long_NGAL_LMM_pvalues_AT_Status_HP.pdf", plot = p_long_NGAL_atn, bg = "transparent", width = 10, height = 8) 

#########---Calprotectin-----------#########################
options(na.action='na.fail')
long_calp.model <- lmer(log2_Calp ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + AD_Status + (1 | ID), data = elisa_long_df, REML = FALSE)
summary(long_calp.model)

#### Adjusted EMMs
long_calp_emm <- pairs(emmeans(long_calp.model, ~ AD_Status), adjust = "BH")

long_calp_ad_comp <- as.data.frame(long_calp_emm)
write.csv(long_calp_ad_comp, "260309_Long_NGAL_AD_LMM_emmeans_results.csv", row.names = F)
long_calp_ad_sig <- long_calp_ad_comp %>%
  mutate(
    label = paste0("q = ", signif(p.value, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast)))

### Plot
p_long_calp_hp <- ggplot(elisa_long_df, aes(x = AD_Status, y = log2_Calp, fill = AD_Status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = AD_Status),shape = 24, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp) +  
  scale_color_manual(values = custom_colors_hp) +
  labs(x = "AD Status",
       y = "Log2 Stool Calprotectin (ug/mL)",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.title = element_text(size = 16),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")    # Prevents clipping of points at the bottom

y_positions <- elisa_long_df %>%
  group_by(AD_Status) %>%
  summarise(y_max = max(log2_Calp, na.rm = TRUE)) %>%
  deframe()

step_height <- 0.5
for(i in seq_len(nrow(long_calp_ad_sig))) {
  g1 <- long_calp_ad_sig$group1[i]
  g2 <- long_calp_ad_sig$group2[i]
  label <- long_calp_ad_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(elisa_long_df$AD_Status) == g1)
  x2 <- which(levels(elisa_long_df$AD_Status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_long_calp_hp <- p_long_calp_hp + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.2, label = label, size = 4.5)
}
ggsave("260309_Long_Calp_LMM_pvalues_AD_Status_HP.pdf", plot = p_long_calp_hp, bg = "transparent", width = 10, height = 8) 

######## Plot by ATN Status ####
options(na.action='na.fail')
long_calp_atn.model <- lmer(log2_Calp ~ Age + Sex + BMI + apoe4_carriage + Diabetes + Hypertension + amyloid_tau_status + (1 | ID), data = elisa_long_df, REML = FALSE)
summary(long_calp_atn.model)

#### Adjusted EMMs
long_calp_emm_atn <- pairs(emmeans(long_calp_atn.model, ~ amyloid_tau_status), adjust = "BH")

long_calp_at_comp <- as.data.frame(long_calp_emm_atn)
write.csv(long_calp_at_comp, "260309_Long_NGAL_AT_LMM_emmeans_results.csv", row.names = F)
long_calp_at_sig <- long_calp_at_comp %>%
  filter(p.value < 0.25) %>%
  mutate(
    label = paste0("q = ", signif(p.value, 2)),
    group1 = gsub("[()]", "", gsub(" -.*", "", contrast)),
    group2 = gsub("[()]", "", gsub(".*- ", "", contrast)))

### Plot
p_long_calp_atn <- ggplot(elisa_long_df, aes(x = amyloid_tau_status, y = log2_Calp, fill = amyloid_tau_status)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = amyloid_tau_status),shape = 24, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_at_status) +  
  scale_color_manual(values = custom_colors_at_status) +
  labs(x = "Amyloid Tau Status",
       y = "Log2 Stool Calprotectin (ug/mL)",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.title = element_text(size = 16),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10)) +  # Increase bottom margin
  coord_cartesian(clip = "off")    # Prevents clipping of points at the bottom

y_positions <- elisa_long_df %>%
  group_by(amyloid_tau_status) %>%
  summarise(y_max = max(log2_Calp, na.rm = TRUE)) %>%
  deframe()

step_height <- 1.2
for(i in seq_len(nrow(long_calp_at_sig))) {
  g1 <- long_calp_at_sig$group1[i]
  g2 <- long_calp_at_sig$group2[i]
  label <- long_calp_at_sig$label[i]
  
  # x-axis positions
  x1 <- which(levels(elisa_long_df$amyloid_tau_status) == g1)
  x2 <- which(levels(elisa_long_df$amyloid_tau_status) == g2)
  
  # y-axis position: a bit above boxplots
  y <- max(y_positions[c(g1,g2)], na.rm = TRUE) + step_height * i
  
  p_long_calp_atn <- p_long_calp_atn + 
    annotate("segment", x = x1, xend = x2, y = y, yend = y) +
    annotate("text", x = mean(c(x1, x2)), y = y + 0.5, label = label, size = 4.5)
}
ggsave("260309_Long_Calp_LMM_pvalues_AT_Status_HP.pdf", plot = p_long_calp_atn, bg = "transparent", width = 10, height = 8) 
```

Figure 4D: Longitudinal GM compositional stability
```{r}
mtg_unif_matrix <- create_pairwise_beta_div_matrix(ps_obj = ps_MTG,
                                                 distance_metric = "unifrac",
                                                 distance_metric_string = "Unifrac distance",
                                                 output_name = "m3_MTG_gt10M_stable_noOther_unif")
mtg_unif_matrix_filt <- mtg_unif_matrix %>%
  filter(Comparison == "Within participant") 

## Only keep pairwse unifrac comparisons from baseline sample
mtg_unif_matrix_baseline <- mtg_unif_matrix_filt %>%
   filter(Time_point.x == 1 | Time_point.y == 1)
# Check which column got the baseline sample. In this case columns y have info 
## on the baseline sample
summary(mtg_unif_matrix_baseline$Time_point.y)

## Select only relevant columns
mtg_unif_matrix_baseline <- mtg_unif_matrix_baseline %>% 
  select(Sample_ID1, Sample_ID2, unifrac, Time_diff_from_baseline_mo.x, Time_diff_from_baseline_mo.y, ID.x, Time_point.x, BMI.x, AD_Status.x, Age.x, Sex.x, apoe4_carriage.x, Diabetes.x, Hypertension.x, ID.y, Time_point.y, BMI.y, AD_Status.y, Age.y, Sex.y, apoe4_carriage.y, Diabetes.y, Hypertension.y) %>%
  dplyr::rename(GM_tax_unif = unifrac)

summary(mtg_unif_matrix_baseline$Time_point.y) 

# Eliminating rows with NAs
mtg_unif_matrix_baseline_subset <- mtg_unif_matrix_baseline %>% filter(complete.cases(.))
summary(mtg_unif_matrix_baseline_subset$AD_Status.y)

# Filter out any consecutive samples that are less than 6 months apart from each other
mtg_unif_matrix_baseline_subset <- mtg_unif_matrix_baseline_subset %>%
  arrange(ID.y, Time_diff_from_baseline_mo.x) %>% 
  group_by(ID.y) %>%
  mutate(
    Time_diff_from_previous_sample_mo = case_when(
      # For the first temporal sample, the previous sample IS the baseline (time = 0)
      # so the diff from previous = diff from baseline
      is.na(lag(Time_diff_from_baseline_mo.x)) ~ Time_diff_from_baseline_mo.x,
      # For all other samples, subtract the previous sample's time from the current
      TRUE ~ Time_diff_from_baseline_mo.x - lag(Time_diff_from_baseline_mo.x)
    )
  ) %>%
  ungroup() %>% filter(Time_diff_from_previous_sample_mo >= 6)

summary(mtg_unif_matrix_baseline_subset$AD_Status.y)
summary(mtg_unif_matrix_baseline_subset$Time_diff_from_baseline_mo.x)

## Time frames more or less match between the two groups
mtg_unif_matrix_baseline_timeframe_hp <- mtg_unif_matrix_baseline_subset %>%
  group_by(AD_Status.y) %>%
  summarise(
    min_timeframe_mo = min(Time_diff_from_baseline_mo.x),
    max_timeframe_mo = max(Time_diff_from_baseline_mo.x))

p_long_unif_timediff_hp <- ggplot(mtg_unif_matrix_baseline_subset, aes(x = Time_diff_from_baseline_mo.x, y = GM_tax_unif, color = AD_Status.y, group = ID.y)) +
  geom_point() +  # Add points
  geom_line(linetype = "dashed") +   # Connect points belonging to the same ID +
  geom_smooth(aes(group = AD_Status.y), method = "lm", se = TRUE) +  # Add a linear regression line with confidence intervals
  scale_color_manual(values = custom_colors_hp) +
  scale_x_continuous(breaks = seq(0, max(mtg_unif_matrix_baseline_subset$Time_diff_from_baseline_mo.x), by = 6)) +  # Set x-axis breaks at every 6 months
  labs(x = "Time difference from baseline sample (Months)", 
       y = "GM Dynamism (UniFrac Dissimilarity)",
       color = "AD Status") +  # Label the axes and legend
  theme_minimal() +  # Use a minimal theme for a clean look
  theme(
    text = element_text(size = 16),       # Set general font size
    axis.title = element_text(size = 18), # Axis titles font size
    axis.text = element_text(size = 14),  # Axis labels font size
    legend.title = element_text(size = 16), # Legend title font size
    legend.text = element_text(size = 14),  # Legend labels font size
    plot.title = element_text(size = 20, face = "bold"), # Optional: Add a title with larger font size
    plot.subtitle = element_text(size = 18))
ggsave("260406_MTG_long_unif_timediff_from_baseline_hp.pdf", plot = p_long_unif_timediff_hp,bg = "transparent", width = 10, height = 8)

# Fit mixed-effects model
options(na.action='na.fail')
lmm_hp.interaction <- lmer(GM_tax_unif ~  Time_diff_from_baseline_mo.x + Age.y + Sex.y + BMI.y + apoe4_carriage.y + Diabetes.y + Hypertension.y + AD_Status.y + Time_diff_from_baseline_mo.x * AD_Status.y + (1 | ID.y), data = mtg_unif_matrix_baseline_subset, REML = FALSE)
summary(lmm_hp.interaction)
lmm_hp_interaction_lmer <- tidy(lmm_hp.interaction)
write.csv(lmm_hp_interaction_lmer, "260308_MTG_long_hp_unif_dynamism_interaction_lmm.csv", row.names = F)

lmm_hp.null <- lmer(GM_tax_unif ~ Time_diff_from_baseline_mo.x + Age.y + Sex.y + BMI.y + apoe4_carriage.y + Diabetes.y + Hypertension.y + (1 | ID.y), data = mtg_unif_matrix_baseline_subset, REML = FALSE)
summary(lmm_hp.null)

lmm_hp.model <- lmer(GM_tax_unif ~ Time_diff_from_baseline_mo.x + Age.y + Sex.y + BMI.y + apoe4_carriage.y + Diabetes.y + Hypertension.y + AD_Status.y + (1 | ID.y), data = mtg_unif_matrix_baseline_subset, REML = FALSE)
summary(lmm_hp.model)
lmm_hp_lmer <- tidy(lmm_hp.model)
write.csv(lmm_hp_lmer, "260308_MTG_long_hp_unif_dynamism_lmm.csv", row.names = F)

anova(lmm_hp.null, lmm_hp.model)

write.csv(anova(lmm_hp.null, lmm_hp.model), "260308_MTG_long_hp_unif_dynamism_lmm_anova.csv")

write.csv(mtg_unif_matrix_baseline_subset, "GM_longitudinal_stability_samples.csv", row.names = F)

```


Figure 4C and Extended Data Figure 5
```{r}
##### Determine time difference between closest and last time point stools 
closest_h_time <- ps_mtg_closest_h_meta %>% select(Sample_ID, ID, Time_diff_from_baseline_mo, AD_Status)
last_h_time <- ps_mtg_last_h_meta %>% select(Sample_ID, ID, Time_diff_from_baseline_mo, AD_Status)

h_sample_pairs <- merge(closest_h_time, last_h_time, by = "ID")


closest_p_time <- ps_mtg_closest_p_meta %>% select(Sample_ID, ID, Time_diff_from_baseline_mo, AD_Status)
last_p_time <- ps_mtg_last_p_meta %>% select(Sample_ID, ID, Time_diff_from_baseline_mo, AD_Status)
p_sample_pairs <- merge(closest_p_time, last_p_time, by = "ID")


hp_sample_pairs <- rbind(h_sample_pairs, p_sample_pairs)

hp_long_network_time_diff <- ggplot(hp_sample_pairs, aes(x = Time_diff_from_baseline_mo.y)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Histogram with Density Curve for HP samples",
       x = "Time difference between 'closest' sample and last follow-up sample (months)",
       y = "Density")
shapiro.test(hp_sample_pairs$Time_diff_from_baseline_mo.y)

hp_sample_pairs_summary <- hp_sample_pairs %>% 
  group_by(AD_Status.x) %>%
  summarise(Q1 = quantile(Time_diff_from_baseline_mo.y, 0.25, na.rm = TRUE),
                                                       Median = median(Time_diff_from_baseline_mo.y, na.rm = TRUE),
                                                       Q3 = quantile(Time_diff_from_baseline_mo.y, 0.75, na.rm = TRUE)) %>%
  ungroup()

hp_sample_pairs_summary2 <- hp_sample_pairs %>% 
  summarise(Q1 = quantile(Time_diff_from_baseline_mo.y, 0.25, na.rm = TRUE),
                                                       Median = median(Time_diff_from_baseline_mo.y, na.rm = TRUE),
                                                       Q3 = quantile(Time_diff_from_baseline_mo.y, 0.75, na.rm = TRUE)) %>%
  ungroup()


## No statistical difference in the time difference from baseline between AD Status groups
wilcox.test(Time_diff_from_baseline_mo.y ~ AD_Status.x, data = hp_sample_pairs)

####### Compare hubbiness metrics between follow-up networks (Fig 5C)

hub_metrics_h_last <- h_last_node %>%
  mutate(Group = "Healthy_follow-up")
hub_metrics_p_last <- p_last_node %>%
  mutate(Group = "Preclinical_follow-up")

hub_metrics_last <- rbind(hub_metrics_h_last, hub_metrics_p_last)
hub_metrics_last$Group <- as.factor(hub_metrics_last$Group)
hub_metrics_last$Group <- factor(hub_metrics_last$Group, levels=c("Healthy_follow-up", "Preclinical_follow-up"))

bp_last_degree <- ggplot(hub_metrics_last, aes(x = Group, y = degree, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp_long) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp_long) +
  labs(x = "AD Status",
       y = "Degree") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy_follow-up", "Preclinical_follow-up")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Fig5C_degree_follow-up.pdf", plot = bp_last_degree, bg = "transparent", width = 10, height = 8)

bp_last_betweenness <- ggplot(hub_metrics_last, aes(x = Group, y = betweenness, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp_long) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp_long) +
  labs(x = "AD Status",
       y = "Betweenness Centrality",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy_follow-up", "Preclinical_follow-up")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 5, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Extended_Fig_5A_betweenness_follow-up.pdf", plot = bp_last_betweenness, bg = "transparent", width = 10, height = 8)

bp_last_closeness <- ggplot(hub_metrics_last, aes(x = Group, y = closeness, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp_long) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp_long) +
  labs(x = "AD Status",
       y = "Closeness Centrality",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")
  ) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy_follow-up", "Preclinical_follow-up")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Extended_Fig_5B_closeness_follow-up.pdf", plot = bp_last_closeness, bg = "transparent", width = 10, height = 8)

bp_last_coreness <- ggplot(hub_metrics_last, aes(x = Group, y = Coreness, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp_long) +  
  scale_color_manual(values = custom_colors_hp_long) +
  labs(x = "AD Status",
       y = "Coreness",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")
  ) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy_follow-up", "Preclinical_follow-up")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Extended_Fig_5C_coreness_follow-up.pdf", plot = bp_last_coreness, bg = "transparent", width = 10, height = 8)

bp_last_efficiency <- ggplot(hub_metrics_last, aes(x = Group, y = local_efficiency, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +  # Avoids plotting outliers as points since we'll overlay scatter
  geom_jitter(aes(color = Group),shape = 21, color = "black", width = 0.2, size = 2, stroke = 0.4, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +  # Adds padding at the bottom
  scale_fill_manual(values = custom_colors_hp_long) +  # Change the palette to suit your preferences
  scale_color_manual(values = custom_colors_hp_long) +
  labs(x = "AD Status",
       y = "Local Efficiency",
       fill = "AD Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18),
        legend.position = "none",
        plot.margin = margin(t = 10, r = 10, b = 50, l = 10),
        axis.line = element_line(color = "black", linewidth = 0.6),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black")
  ) +  # Increase bottom margin
  coord_cartesian(clip = "off") +  # Prevents clipping of points at the bottom
  stat_compare_means(comparisons = list(c("Healthy_follow-up", "Preclinical_follow-up")),
                     method = "wilcox.test", p.adjust.method = "BH", label = "p.format",
                     step.increase = 0.1, # Helps prevent label overlap
                     size = 6, #adjusted p-value label size
                     show.legend = FALSE) 
ggsave("Extended_Fig_5D_local_effiency_follow-up.pdf", plot = bp_last_efficiency, bg = "transparent", width = 10, height = 8)
```

Figure 5E: Longitudinal guild abundance stability
```{r}
library(lme4)
library(lmerTest)
long_guild_metadata <- read_excel("~/Documents/GitHub/PreclinicalAD_GutMicrobiome-Inflammation/R/Data/SupplementaryTable1_final.xlsx")

stability_cohort <- read_excel("~/Documents/GitHub/PreclinicalAD_GutMicrobiome-Inflammation/R/Data/GM_longitudinal_stability_samples.xlsx")
stability_cohort_ids <- unique(union(stability_cohort$Sample_ID1, stability_cohort$Sample_ID2))

long_guild_meta_filt <- long_guild_metadata %>% filter(Sample_ID %in% stability_cohort_ids)
long_guild_meta_filt$Diabetes <- as.factor(long_guild_meta_filt$Diabetes)
long_guild_meta_filt$Hypertension <- as.factor(long_guild_meta_filt$Hypertension)
options(na.action='na.fail')
columns_to_check <- c("guild_abundance", "AD_Status", "Sex", "Age", "BMI", "Diabetes", "Hypertension", "apoe4_carriage", "Time_diff_from_baseline_mo")
long_guild_meta_filt <- long_guild_meta_filt %>%
  filter(if_all(all_of(columns_to_check), ~ !is.na(.)))

model <- lmer(guild_abundance ~ AD_Status+ Sex + Age + BMI+ Diabetes+ Hypertension + apoe4_carriage+ Time_diff_from_baseline_mo + (1|ID), data = long_guild_meta_filt)
null_model <- lmer(guild_abundance ~ Sex + Age + BMI+ Diabetes+ Hypertension + apoe4_carriage+ Time_diff_from_baseline_mo + (1|ID), data = long_guild_meta_filt)


# View results
summary(model)

# Get p-values
anova(model, null_model) 
write.csv(anova(model, null_model), "~/Downloads/260406_MTG_long_guild_abund_hp_lmm_anova.csv")
write.csv(as.data.frame(summary(model)$coefficients), "~/Downloads/260406_MTG_long_guild_abund_time_from_baseline_lmm.csv")

p_long_guild_timediff_hp <- ggplot(long_guild_meta_filt, aes(x = Time_diff_from_baseline_mo, y = guild_abundance, color = AD_Status, group = ID)) +
  geom_point() +  # Add points
  geom_line(linetype = "dashed") +   # Connect points belonging to the same ID +
  geom_smooth(aes(group = AD_Status), method = "lm", se = TRUE) +  # Add a linear regression line with confidence intervals
  scale_color_manual(values = custom_colors_hp) +
  scale_x_continuous(breaks = seq(0, max(long_guild_meta_filt$time_from_baseline_mo), by = 6)) +  # Set x-axis breaks at every 6 months
  labs(x = "Time difference from baseline sample (Months)", 
       y = "Guild abundance",
       color = "AD Status") +  # Label the axes and legend
  theme_minimal() +  # Use a minimal theme for a clean look
  theme(
    text = element_text(size = 16),       # Set general font size
    axis.title = element_text(size = 18), # Axis titles font size
    axis.text = element_text(size = 14),  # Axis labels font size
    legend.title = element_text(size = 16), # Legend title font size
    legend.text = element_text(size = 14),  # Legend labels font size
    plot.title = element_text(size = 20, face = "bold"), # Optional: Add a title with larger font size
    plot.subtitle = element_text(size = 18) # Optional: Add subtitle with a larger font size
  )
ggsave("~/Documents/260401_MTG_long_guild_abund_timediff_from_baseline_hp.pdf", plot = p_long_guild_timediff_hp,bg = "transparent", width = 10, height = 8)


```



